---
title: "DwitiBagadia_A03_DataExploration"
author: "Dwiti Bagadia"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r working directory and loading packages, message=FALSE}
#load packages and set working directory
library(tidyverse)
library(lubridate)
getwd()
setwd('/Users/d/Desktop/UNC/Spring:23/DUKE 872 L1 - Environmental Data Analysis/EDA-Spring23/Data/Raw')

#upload datasets
Neonics <- read.csv("ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Since the 1990s, neonicotinoids have become one of the most commonly used insecticides in the U.S. and the world. They are now widely used in the agricultural sector as they possess a low mammalian toxicity and are effective for controlling stubborn pests in the soil. As neonicotinoids have proved to lead to loss of pollinators, they can give us a better understanding on the mechanisms behind the various diseases caused to the pollinators. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and the woody debris may act as a tinder that encourages the spreading of forest fires and in some cases also the start of the forest fire; in addition to their negative impacts they even act as a habitat for insects. Thus understanding about the quantity of the habitat for insects and influence on carbon budgets is important for integrating the ecotoxicity data. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected from elevated and ground traps. 
 2. The mass data is collected to an accuracy of 0.01 grams, for each group.
 3. The randomness of trap placements is based on the type of vegetation.  

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r NEONICS Summary}
#obtaining dimensions of NEONICS dataset
dim(Neonics)
```

> Answer: 4623(R). 30(C)

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r summary of effect column in Neonics}
#Summarizing the effects column
summary(Neonics$Effect)
```

> Answer: The most common effects of Neonics(Population and Mortality) help us understand about how common mortality is a result of neonicotinoids. It helps us by indicating the non-selective nature of the insecticide and the serious effect it has on the pollinators. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r commonly studied species - summary}
#creating a sorted summary of if the Species with common name
sort(summary(Neonics$Species.Common.Name), decreasing = TRUE, na.last = TRUE)
```

> Answer: The most commonly studied insects were - Honey Bee (667) and Parasitic Wasp (285). The honey bees being one of the most common insects which benfit from pollinators is the one that is studied the most. As Neonicotinoids are extremely dangerous to pollinators and benefitting insects, this data is important to study.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r checking class of Conc.1..Author }
#class function to check the data type
class("Conc.1..Author")
```

> Answer: The Conc.1..Author is a character class, as the column has some data with both numbers and '<' signs. When some data in the column are charaters, the class of the vector wukk convert to character instead of numeric.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r plot studies by publication year}
#creating a plot of number of studies done by publication year
ggplot(Neonics) + geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r adding color to the number of studies x publication year graph}
#set of commands to plot the graph with colors
is.na(Neonics$Test.Location)
NeonicsComplete <- na.omit(Neonics$Test.Location)
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Lab is the most test location more than once, first in 2000-2005 and then again in 2010-2020. While in time span of 1990-2000 and 2005-2010 the natural field was more common test location. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r graph for Endpoint Counts}
#plotting endpoint counts
ggplot(Neonics) + 
  geom_bar(aes(x = Endpoint)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

> Answer: LOEL and NOEL are the most common end points. LOEL is the lowest observable effect level - having the lowest dose of insecticides producing effects that were significantly different from responses of controls. NOEL is the no observable effect level - gavubg the highest dose producing effects that were not significantly different from the responses of controls according to the statistical tests reported by the author. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r determining the class of CollectDate and converting it to date if not}
#data class check
class(Litter$collectDate)
#output is factor#converting to date 
Litter$collectDate <- as.Date(Litter$collectDate, format("%y/%m/%d"))
#checking class
class(Litter$collectDate)
#summary and check
summary(Litter$collectDate <- as.Date(Litter$collectDate, format = "%y%m%d"))
unique(Litter$collectDate)
#i cannot figure out what went wrong here. 
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r summary vs. unique}
#unique for Niwot Ridge
unique(Litter$siteID)
summary(Litter$siteID)
```

> Answer: 188 plots were sampled at Niwot Ridge, unique command eliminated all the duplicate values, so the outcome was only 1 value - NIWO. where the summary command calculates the number of each value hence, giving us the number of the total plots (188) sampled at Niwot Ridge

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r functional groups bar graph}
ggplot(Litter) + geom_bar(aes(x=functionalGroup)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r box plotfunctional group}
#box plot 
ggplot(Litter) + geom_boxplot(aes(x=functionalGroup, y=dryMass)) + coord_flip()
```
```{r}
#violin plot
ggplot(Litter) + geom_violin(aes(x=functionalGroup, y=dryMass)) + coord_flip()
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: In this case, Ieach functional group more aesthetically and captures the viewers attention to each group. The box plot shows only the summarized data, where as the violin plot shows all the data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needless 
