---
title: "Assignment 10: Data Scraping"
author: "Dwiti Bagadia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, message = FALSE, warning = FALSE}
#1 loading packages
library(tidyverse)
library(lubridate)
library(formatR)
library(rvest)
library(ggplot2)

#working directory
getwd()

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
the.web = read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- the.web %>% 
  html_nodes("div + table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  
PWSID <- the.web %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
  
ownership <- the.web %>%
  html_nodes("div + table tr:nth-child(2) td:nth-child(4)") %>% html_text()

max.withdrawals.mgd <- the.web %>%
  html_node("th ~ td + td") %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the average daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4
withdrawal <- data.frame("Month" = rep(1:12),"Water System Name"= water.system.name, "Ownership" = ownership, "PWSID" = PWSID, "MGD" = as.numeric(max.withdrawals.mgd))
withdrawal <- withdrawal %>% mutate(Date = ym(paste0("2022", "-", Month)))

#5
MGD.plot <- ggplot(withdrawal) + geom_line(aes(x = Date, y = MGD)) + 
  ylab("Max Day Use(Gallons)") + xlab("Date") + 
  labs(title = "Max Daily Withdrawals Across the Months (2022) in Durham") + 
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b")
print(MGD.plot)

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(pwsid, the_year){the_url = ifelse(pwsid == "03-32-010" & the_year == 2022, "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022", paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php", "?", "pwsid=", pwsid, "&", "year=", the_year))
web <- read_html(the_url)
water.system.name.2 <- web %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
PWSID.2 <- web %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
ownership.2 <- web %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
max.withdrawals.mgd.2 <- web %>% 
  html_nodes("th~ td+ td") %>% html_text()
withdrawal.2 <- data.frame("Month" = rep(1:12),"Water System Name"= water.system.name.2, "Ownership" =  ownership.2, "PWSID" = PWSID.2, "MGD" = as.numeric(max.withdrawals.mgd.2))
withdrawal.2 <- withdrawal.2 %>% 
  mutate(Date = ym(paste0(the_year, "-", Month)))
return(withdrawal.2)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
Durham.2015 <- scrape.it("03-32-010", 2015)
Durham.2015.plot <- ggplot(Durham.2015) + geom_line(aes(x = Date, y = MGD))+ 
  ylab("Max Day Use(Gallons)") + xlab("Date") + 
  labs(title = "Max Daily Withdrawals Across the Months for 2015 in Durham") + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
print(Durham.2015.plot)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
Asheville.2015 = scrape.it("01-11-010", 2015)
Asheville.2015
Compare.plot = ggplot(Durham.2015, aes(x = Date, y = MGD)) + geom_line(color = "lightblue")+ 
  ylab("Max Day Use(Gallons)") + xlab("Date") + 
  geom_line(data = Asheville.2015, color = "lightpink") + 
  labs(title = "Max Daily Withdrawals Across the Months for 2015 in Durham and Asheville") + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
print(Compare.plot)

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
Asheville.years = map2("01-11-010", c(2010:2021), scrape.it)
Ashe.years.df = bind_rows(Asheville.years)
Ashe.years.plot = ggplot(Ashe.years.df, aes(x = Date, y = MGD)) + 
  geom_line() + geom_smooth(method = "loess") + 
  ylab("Max Day Use(Gallons)") + xlab("Date") + 
  labs(title = "Max Daily Withdrawals Across the Months for Asheville (2010-2021)") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
print(Ashe.years.plot)

```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
We see a trend in water usage in Asheville just by looking at the plot. We see a slow smooth decrease in usage from 2010 - 2015 and then a significant increase in usage from 2016 - 2021.