---
title: "Assignment 2: Coding Basics"
author: "Dwiti Bagadia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. generating sequence
seq(1,100)

#2. mean and median of the sequence and renaming the sequence
#setting a2 = 100
a2 <- seq(1,100)

#mean 
mean(a2)
mean_a2 <- mean(a2)
mean_a2

#median
median(a2)
median_a2 <- median(a2)
median_a2

#3.comparison
mean_a2 > median_a2

#mean and median are equal 
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5 
#name of students 
student_names <- c("Dwiti, Ria, Aditi, Sanjana")
student_names #character

#test scores 
test_scores <- c(80, 85, 90, 49)
test_scores #numeric

#pass/fail condition
fail <- ifelse(test_scores<50, TRUE, FALSE)
fail #logical 
```
```{r basics 2.1}
#6 
class(student_names)
class(test_scores)
class(fail)
```
```{r basics 2.2}
#7
df_studentdata <- data.frame(student_names,test_scores,fail)
df_studentdata
```
```{r}
names(df_studentdata) <- c('Name', 'Score', 'Fail')
df_studentdata
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Where matrix contain only a single type of data, data frames can contain various types of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}
#10 
```


11. Apply your function to the vector with test scores that you created in number 5.

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
